---
title: "Strings: In-class Exercises (Part 2)"
format:
  pdf: default
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

You can download this .qmd file from
[here](https://github.com/joeroith/264_spring_2026/blob/main/03_strings_part2.qmd).
Just hit the Download Raw File button.

This uses parts of R4DS Ch 14: Strings and Ch 15: Regular Expressions
(both the first and second editions).

## Manipulating strings

str functions to know for manipulating strings:

-   str_length()
-   str_sub()
-   str_c()
-   str_to_lower()
-   str_to_upper()
-   str_to_title()
-   str_replace() *not in 02_strings examples*

```{r}
#| message: false
#| warning: false

library(tidyverse)

#spotify <- read_csv("Data/spotify.csv") 
spotify <- read_csv("https://joeroith.github.io/264_spring_2026/Data/spotify.csv")

spot_smaller <- spotify |>
  select(
    title, 
    artist, 
    album_release_date, 
    album_name, 
    subgenre, 
    playlist_name
  )

spot_smaller <- spot_smaller[c(5, 32, 49, 52, 83, 175, 219, 231, 246, 265), ]
spot_smaller
```

## Warm-up

0.  Describe what EACH of the str\_ functions below does. Then, create a
    new variable "month" which is the two digit month from
    album_release_date

```{r}
spot_new <- spot_smaller |>
  select(title, album_release_date) |>
  mutate(title_length = str_length(title), # new column with length of title
         
         year = str_sub(album_release_date, 1, 4), # new column year with just the year
         
         month = parse_number(str_sub(album_release_date, 6,7)),
         
         #make letters lowercase in title
         title_lower = str_to_lower(title),
         
         album_release_date2 = str_replace_all(album_release_date, "-", "/"))
         # Will create a new column where the date will have / instead of -

spot_new
         
max_length <- max(spot_new$title_length)

#If we want to combine text and numbers/ variables, in this case we want to combine the mas length that we find above
str_c("The longest title is", max_length, "characters long.", sep = " ")
str_glue("The longest title is {max_length} characters long.")
```

## Important functions for identifying strings which match

str_view() : most useful for testing str_subset() : useful for printing
matches to the console str_detect() : useful when working within a
tibble

1.  Identify the input type and output type for each of these examples:

```{r}

#str_view(spot_smaller$subgenre, "pop") #or

#str_view(spot_smaller$subgenre, "pop", match = NA) 
# to see the whole list even those that don't match
  #input: vector of characters
  #output: stringr_view class

typeof(str_view(spot_smaller$subgenre, "pop"))
class(str_view(spot_smaller$subgenre, "pop"))

class(str_view(spot_smaller$subgenre, "pop", match = NA))
  #input: vector of characters
  #output: stringr_view class

#str_view(spot_smaller$subgenre, "pop", html = TRUE)

str_subset(spot_smaller$subgenre, "pop")
  #input vector 
  #output vector

str_detect(spot_smaller$subgenre, "pop")
  #input : string vector 
  #output: logical vector
```

2.  Use str_detect to print the rows of the spot_smaller tibble
    containing songs that have "pop" in the subgenre. (i.e. make a new
    tibble with fewer rows)
    
```{r}
(pop_small <- spot_smaller |>
  filter(str_detect(spot_smaller$subgenre, "pop"))|>
  select(title, subgenre))

# true or false
```
    

3.  Find the mean song title length for songs with "pop" in the subgenre
    and songs without "pop" in the subgenre.
    
```{r}
(pop_mean_length <- spot_smaller |>
  mutate(
    sub_pop = ifelse(str_detect(spot_smaller$subgenre,"pop"), "Genre with pop", "Genre without pop")) |>
  group_by(sub_pop) |>
  summarize(mean_title_length = mean(str_length(title))))

pop_mean_length

```


Producing a table like this would be great:

```{r}
#| eval: false

# A tibble: 2 × 2
 # sub_pop mean_title_length
  #<lgl>               <dbl>
#1 FALSE                18.6
#2 TRUE                 13.6
```

Producing a table like this would be SUPER great (hint: `ifelse()`):

```{r}
#| eval: false

# A tibble: 2 × 2
  #sub_pop           mean_title_length
 # <chr>                         <dbl>
#1 Genre with pop                 13.6
#2 Genre without pop              18.6
```

4.  In the bigspotify dataset, find the proportion of songs which
    contain "love" in the title (track_name) by playlist_genre.

```{r}
#| message: false
#| warning: false

bigspotify <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-21/spotify_songs.csv')

bigspotify
```

```{r}

#First
#str_view(bigspotify$track_name, "love")
#str_view(str_to_lower(bigspotify$track_name, "love"))

(lowerbigspotify <- bigspotify |>
  #Get ride of NA names
  filter(!is.na(track_name)) |>
  
  #to create new variables
  mutate(
    #Make sure all names are in lower case so we can look for love
    lower_track_name = str_to_lower(track_name),
    # new column that will say true or false if love appears
    love = str_detect(lower_track_name, "love")) |>
    
  #Group by genre of playlist
  group_by(playlist_genre) |>
  #will summarize and create new column that takes the mean of love
  summarize(prop_love = mean(love)))

```



## Matching patterns with regular expressions

|          |                                          |
|----------|------------------------------------------|
| `^abc`   | string starts with exactly "abc"         |
| `abc$`   | string ends with exactly "abc"           |
| `.`      | any character (helpful as a placeholder) |
| `[^abc]` | anything EXCEPT a or b or c              |

```{r}
# Guess the output!

#start with K
#str_view(spot_smaller$artist, "^K")


#Songs that are release on the frist day of the month
#str_view(spot_smaller$album_release_date, "01$")

#starts with any two chracters and then a space
#str_view(spot_smaller$title, "^.. ")

#Artists that dont have any letters in it
#str_view(spot_smaller$artist, "[^A-Za-z ]")
```

5.  Given the corpus of common words in stringr::words, create regular
    expressions that find all words that:

-   Start with “y”.
-   End with “x”
-   Are exactly three letters long.
-   Have seven letters or more.
-   Start with a vowel.
-   End with ed, but not with eed.
-   Words where q is not followed by u. (are there any in `words`?)

```{r}
# Try using str_view() or str_subset()

str_subset(stringr::words,"^y")

str_subset(stringr::words,"x$")

str_subset(stringr::words,"^...$") 

str_subset(stringr::words,"^.{7,}$") 

str_subset(stringr::words,"^[aeiou]") 

str_subset(stringr::words,"[^e]ed$")

str_subset(stringr::words,"q[^u]")


# For example, to find words with "tion" at any point, I could use:
#str_view(words, "tion")
str_subset(words, "tion")
```

## More useful regular expressions:

|       |                              |
|-------|------------------------------|
| `\\d` | any number                   |
| `\\s` | any space, tab, etc.         |
| `\\b` | any boundary: space, ., etc.
| Use capital D, S , B to negate it

```{r}

#Any digit
#str_view(spot_smaller$album_name, "\\d")

#Any space/tab/etc.
#str_view(spot_smaller$album_name, "\\s")

#Any boundary/ break or commas
#str_view(spot_smaller$album_name, "\\b")
```

Here are the regular expression special characters that require an
escape character (a preceding `\`): `^` `$` `.` `?` `*` `|` `+` `(` `)`
`[` `{`

For any characters with special properties, use `\` to "escape" its
special meaning ... but `\` is itself a special character ... so we need
two `\\`! (e.g. `\\$`, `\\.`, etc.)

```{r}
#str_view(spot_smaller$title, "$")
#str_view(spot_smaller$title, "\\$")
```

6.  In bigspotify, how many track_names include a `$`? Be sure you print
    the track_names you find and make sure the dollar sign is not just
    in a featured artist!
    
```{r}
bigspotify |>
  filter(str_detect(track_name,"\\$")) |>
  select(track_name, track_artist)

# Detect if you use it in a filter and want a tibble
# in this case it is giving us the artist names in the track name that have $ and not just the track name so we do the following 


bigspotify |>
  filter(str_detect(track_name,"\\$")) |>
  filter(!str_detect(track_name, "(feat|with).*\\$")) |>
  select(track_name, track_artist)


```

7.  In bigspotify, how many track_names include a dollar amount (a `$`
    followed by a number).
```{r}   
#Result 
# A tibble: 2 × 2
  #track_name                  track_artist
  #<chr>                       <chr>       
#1 $20 Fine                    Jimi Hendrix
#2 $100 (feat. Polo Donatello) Mibbs  
```
    
```{r}
bigspotify |>
  filter(str_detect(track_name, "\\$\\d")) |>
  select(track_name, track_artist)
  
```

## Repetition

|         |                       |
|---------|-----------------------|
| `?`     | 0 or 1 times          |
| `+`     | 1 or more             |
| `*`     | 0 or more             |
| `{n}`   | exactly n times       |
| `{n,}`  | n or more times       |
| `{,m}`  | at most m times       |
| `{n,m}` | between n and m times |

```{r}

# Any capital letter that are back to back A_Z
#str_view(spot_smaller$album_name, "[A-Z]{2,}")

#str_view(spot_smaller$album_release_date, "\\d{4}-\\d{2}")
```

**Use at least 1 repetition symbol when solving 8-10 below**

8.  Modify the first regular expression above to also pick up "A.A" (in
    addition to "BEYONC" and "II"). That is, pick up strings where there
    might be a period between capital letters.

```{r}

#str_view(spot_smaller$album_name, "[A-Z.]{2,}")

#str_view(spot_smaller$album_name, "([A-Z]\\.?){2,}")

#str_view(spot_smaller$album_name, "[A-Z]\\.?[A-Z]")

```

9.  Create some strings that satisfy these regular expressions and
    explain.

-   `"^.*$"`, decimal meaning anything, starts with(^) any character(.) one(*) or more times and end ($) with any character

-   `"\\{.+\\}"`, escaping curly braces so we are looking for curly braces (\\{}), decimal(.) looking for anything and (+) one or more times , and ends with curly braces (\\})

```{r}
#Starts with any character 
#str_view(c("text"),"^.*$" )

#Anything that repeats 0 or 1 or more times 
#str_view(c("{text}","text", "{}", "{ggplot}"),"\\{.+\\}")

```

10. Create regular expressions to find all `stringr::words` that:

-   Start with three consonants. Starts with (^) and be a constant [^aeiou] and exactly three times {3}
-   Have two or more vowel-consonant pairs in a row.

```{r}

#str_view(stringr::words,"^[^aeiou]{3}")

#str_view(stringr::words,"([aeiou][^aeiou]|[^aeiou][aeiou]){2,}" )

#str_view(stringr::words,"([aeiou][^aeiou]){2,}" )

```

## Useful functions for handling patterns

str_extract() : extract a string that matches a pattern str_count() :
count how many times a pattern occurs within a string

```{r}
str_extract(spot_smaller$album_release_date, "\\d{4}-\\d{2}")

spot_smaller |>
  select(album_release_date) |>
  mutate(year_month = str_extract(album_release_date, "\\d{4}-\\d{2}"))

spot_smaller |>
  select(artist) |>
  mutate(n_vowels = str_count(artist, "[aeiou]"))
```

11. In the spot_smaller dataset, how many words are in each title?
    (hint: `\\b`)
    
```{r}

spot_smaller |>
  mutate(word_count = str_count(title, "\\b\\w+\\b")) |>
  select(title, word_count)
  
```

12. In the spot_smaller dataset, extract the first word from every
    title. Show how you would print out these words as a vector and how
    you would create a new column on the spot_smaller tibble. That is,
    produce this:

```{r}
# [1] "Hear"      "Run"       "Formation" "7/11"      "My"        "It's"     
# [7] "Poetic"    "A.D.H.D"   "Ya"        "Runnin"   
```

```{r}

str_extract(spot_smaller$title, "^\\S+")

```

Then this:

```{r}
# A tibble: 10 × 2
#   title                                             first_word
#   <chr>                                             <chr>     
# 1 Hear Me Now                                       Hear      
# 2 Run the World (Girls)                             Run       
# 3 Formation                                         Formation 
# 4 7/11                                              7/11      
# 5 My Oh My (feat. DaBaby)                           My        
# 6 It's Automatic                                    It's      
# 7 Poetic Justice                                    Poetic    
# 8 A.D.H.D                                           A.D.H.D   
# 9 Ya Estuvo                                         Ya        
#10 Runnin (with A$AP Rocky, A$AP Ferg & Nicki Minaj) Runnin    
```

```{r}
spot_smaller |>
  mutate(first_word = str_extract(spot_smaller$title, "^\\S+")) |>
           select(title, first_word)

```

13. Which decades are popular for playlist_names? Using the bigspotify
    dataset, try doing each of these steps one at a time!

-   filter the bigspotify dataset to only include playlists that include
    something like "80's" or "00's" in their title.
-   create a new column that extracts the decade
-   use count to find how many playlists include each decade
-   what if you include both "80's" and "80s"?
-   how can you count "80's" and "80s" together in your final tibble?

```{r}
bigspotify |>
  filter(str_detect(playlist_name, "\\b\\d{2}'s\\b")) |>
  mutate(decade = str_extract(playlist_name, "\\b\\d{2}'?s\\b"),
         decade = str_replace(decade, "'", "")) |>
  count(decade)
```

## Grouping and backreferences

```{r}
# find all fruits with repeated pair of letters.  
fruit = stringr::fruit
fruit
#str_view(fruit, "(..)\\1", match = TRUE)
#If i see those same two characters repeated detect those, .. and do i see it again so For example la and do I see it again la so balala would show up 

#Result 
#[4] │ b<anan>a
#[20] │ <coco>nut
#[#22] │ <cucu>mber
#[41] │ <juju>be
#[56] │ <papa>ya
#[73] │ s<alal> berry


# why does the code below add "pepper" and even "nectarine"?
#str_view(fruit, "(..)(.*)\\1", match = TRUE)
  # So two characters, then any characters and then the first ones will repeat
```

Tips with backreference:

-   You must use () around the the thing you want to reference.
-   To backreference multiple times, use `\\1` again.
-   The number refers to which spot you are referencing... e.g. `\\2`
    references the second set of ()

```{r}
x1 <- c("abxyba", "abccba", "xyaayx", "abxyab", "abcabc")
#str_view(x1, "(.)(.)(..)\\2\\1")
#(first position) (second position)
#(a) (b) (cc) \\2nd position \\1 position
#str_view(x1, "(.)(.)(..)\\1\\2")
#str_view(x1, "(.)(.)(.)\\1\\2\\3")
```

14. Describe to your groupmates what these expressions will match, and
    provide a word or expression as an example:

-   `"(.)\\1\\1"` Will find words that repeat a character three times in  a row

-   `"(.)(.)(.).*\\3\\2\\1"` Will find words that have  3 repeating characters in reverse order after skipping the 4th one

Which words in `stringr::words` match each expression?

-   `"(.)\\1\\1"`  : no words match
```{r}
#str_view(stringr::words, "(.)\\1\\1")
```

-   `"(.)(.)(.).*\\3\\2\\1"` : paragraph
```{r}
#str_view(stringr::words, "(.)(.)(.).*\\3\\2\\1")
```

15. Construct a regular expression to match words in `stringr::words`
    that contain a repeated pair of letters (e.g. “church” contains “ch”
    repeated twice) but *not* match repeated pairs of numbers (e.g.
    507-786-3861).
```{r}
#str_view(stringr::words, "([A-Za-z]{2}).*\\1")
```

16. Reformat the album_release_date variable in spot_smaller so that it
    is MM-DD-YYYY instead of YYYY-MM-DD. (Hint: str_replace().)
```{r}

spot_smaller |>
  mutate(
    album_release_date = str_replace(album_release_date, 
                                     "(\\d{4})-(\\d{2})-(\\d{2})","\\2-\\3-\\1")) |>
  select(title, album_release_date)
```

17. BEFORE RUNNING IT, explain to your partner(s) what the following R
    chunk will do:
    
    The following r chunk will replace the second and third words positions, and will only hsow us the first 5.

```{r}
sentences |> 
  str_replace("([^ ]+) ([^ ]+) ([^ ]+)", "\\1 \\3 \\2") |> 
  head(5)
```